---
title: 'APSTA 2047: HW1'
author: "William Spagnola & Emanuel Bazov"
date: "9/18/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)



'
fares <- read_csv("trip_fare_8.csv")
trips <- read_csv("trip_data_8.csv")
'

```

```{r, include = FALSE}
'
trips_hw_1  <- trips %>% filter(pickup_datetime >= ymd_hms("2013-08-15 00:00:00") 
                 & pickup_datetime < ymd_hms("2013-08-16 00:00:00"))

fares_hw_1 <- fares %>% filter(pickup_datetime >= ymd_hms("2013-08-15 00:00:00") 
                 & pickup_datetime < ymd_hms("2013-08-16 00:00:00"))


trips_hw_1 %>% arrange(desc(pickup_datetime))
trips_hw_1 %>% arrange(pickup_datetime)

fares_hw_1 %>% arrange(desc(pickup_datetime))
fares_hw_1 %>% arrange(pickup_datetime)


write_csv(x = trips_hw_1, path = "trips_hw_1.csv")
write_csv(x = fares_hw_1, path = "fares_hw_1.csv")
'
```


#Clean Data
```{r Clean}

trips_hw_1 <- read_csv("trips_hw_1.csv")



summary(trips_hw_1)


#Look at Trip Time 
trips_hw_1 %>%
      arrange(trip_time_in_secs)

#Filter Trips less than 60 seconds
trips_hw_1 <- trips_hw_1 %>% filter(trip_time_in_secs > 60)

#Generate column of trip time minutes
trips_hw_1 <- trips_hw_1 %>%
  mutate(trip_time_in_mins =  (trip_time_in_secs / 60)) 


#Generate column of trip time Hours
trips_hw_1 <- trips_hw_1 %>%
  mutate(trip_time_in_hrs =  (trip_time_in_mins / 60)) 

#Look at Max Trip Time in descending Order
trips_hw_1 %>%
      arrange(desc(trip_time_in_mins))
trips_hw_1 %>%
      arrange(desc(trip_time_in_hrs))

#Filter Trip Time to be less than or equal to 3 hours 
trips_hw_1 <- trips_hw_1 %>% 
  filter(trip_time_in_hrs  <= 3)
    
#Generate column of average speed
trips_hw_1 <- trips_hw_1 %>%
  mutate(avg_speed = trip_distance / (trip_time_in_secs / 3600)) 

#Examing Average Speed
summary(trips_hw_1$avg_speed)

#Filter Average Speed to be Less than or equal to 60
trips_hw_1 <- trips_hw_1 %>%
              filter(avg_speed <= 60)

#Filter Average Speed to be greater than or to 10
trips_hw_1 <- trips_hw_1 %>%
              filter(avg_speed >= 10)




```


##Clean Data based on Workday Length
```{r}
#Calculate Total Numbwer of Drivers
length(unique(trips_hw_1$hack_license))

#Check Shortest Work Days
trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  arrange(workday_in_seconds)


#Check Longest Work Days
trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  arrange(desc(workday_in_seconds))

#Look at people who worked more than 16 hours
trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  filter(workday_in_seconds > as.duration(16*3600) )
  
#Look at people who worked less than 1 hour
trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  filter(workday_in_seconds < as.duration(3600) )       

#Create Boxplot of Distribution of Hack licenses
trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  ggplot(aes(x=1, y= as.numeric(workday_in_seconds)/3600))  + geom_boxplot()

#Pull License Numbers of Cab Driver Who Worked at least one Hour   
at_least_1hr <- trips_hw_1 %>%
  group_by(hack_license) %>%
  summarize(earliest_start= min(pickup_datetime), latest_finish = max(dropoff_datetime),
            workday_in_seconds = as.duration(latest_finish- earliest_start) ) %>%
  filter(workday_in_seconds >= as.duration(3600) ) %>%
  pull(hack_license)


#Calculate Rows Deleted
nrow(trips_hw_1) -trips_hw_1 %>%
    filter(hack_license %in% at_least_1hr ) %>%
  nrow()

#Remove Cab Drivers who Worked less than 1 hours
trips_hw_1 <- trips_hw_1 %>%
    filter(hack_license %in% at_least_1hr )
```


#Compare Drop Off, Pick Up Date, Trip time in Sec
```{r}

#Remove Rows Where Difference between Dropoff - Pickup and trip_time in secs is greater than 1  minute
trips_hw_1 <- trips_hw_1 %>% 
mutate(trip_duration = as.duration(dropoff_datetime - pickup_datetime) ) %>%
    mutate(trip_diff = abs(trip_time_in_secs - as.numeric(trip_duration)) ) %>%
    filter(trip_diff<= 60)

```


###Check Passenger Count
```{r}
summary(trips_hw_1$passenger_count)

```
**How many people can fit into a yellow taxicab?**

*From Driver Rule 54-15(g) (in PDF): The maximum amount of passengers allowed in a yellow taxicab by law is four (4) in a four (4) passenger taxicab or five (5) passengers in a five (5) passenger taxicab, except that an additional passenger must be accepted if such passenger is under the age of seven (7) and is held on the lap of an adult passenger seated in the rear.*

SOURCE(http://www.nyc.gov/html/tlc/html/faq/faq_pass.shtml)


```{r Cleaning Fares Data}
fares_hw_1 <- read_csv("fares_hw_1.csv")

#Examining fare variables to see amounts that are unreasonable
summary(fares_hw_1)

#filter fare amounts less than $2.50 and unreasonably high fares. 
fares_hw_1<- fares_hw_1 %>%
  filter(fare_amount>2.5 & fare_amount<=200)

#Reexamining fare amounts. Looks Ok. 
summary(fares_hw_1$fare_amount)

#Boxplot to check surcharge distribution
boxplot(fares_hw_1$surcharge)

#filter out surcharges that are negative and too high
fares_hw_1<- fares_hw_1 %>%
   filter(surcharge>=0 & surcharge<367.46)

#Filter out 0s for MTA Tax
fares_hw_1<-fares_hw_1 %>%
  filter(mta_tax>0)

#Filter out tips greater than $100 and by tip amounts and imputed tip that differ by more than $1 
fares_hw_1<-fares_hw_1 %>%
  filter(tip_amount<100) %>% 
  mutate(imputed_tip=total_amount-(fare_amount+surcharge+mta_tax+tip_amount+tolls_amount))
fares_hw_1<-fares_hw_1 %>% 
  mutate(tip_diff = abs(imputed_tip - tip_amount), tip_diff_amount=abs(tip_diff-tip_amount)) %>% 
  filter(tip_diff_amount<1)

#Drop irrelevant columns
fares_hw_1<-fares_hw_1 %>%
  select(-tip_diff, -tip_diff_amount, -imputed_tip)


#Cleaning tolls by getting rid of amounts that are too high
fares_hw_1<- fares_hw_1%>%
  filter(tolls_amount<60)

#See how everything looks
summary(fares_hw_1)
```


## Part D. Merge Trips and Fares into One Dataframe called Taxi Data
```{r}
taxi_data <- full_join(x = trips_hw_1, y = fares_hw_1) 
(nrow(taxi_data))
taxi_data <- distinct(taxi_data) #Remove Duplicate Rows
(nrow(taxi_data))
taxi_data<- taxi_data %>% select(-store_and_fwd_flag) #Remove column with missing data. 
taxi_data<- na.omit(taxi_data)  #Remove all cases with missing data. 
(nrow(taxi_data))
```




## Part E. Summary Statistics for Each Taxi Cab
```{r}
taxi_data  %>% 
    group_by(medallion) %>%
    summarize(total_trips = n(),
              total_passengers  = sum(passenger_count),
              total_time_with_passengers = sum(trip_time_in_hrs),
              total_distance = sum(trip_distance),
              total_earnings=sum(total_amount))
#NOTE! : Add Earnings once we Merge with fares
    
```


##Part F
```{r}


#Extract hours from "pickup_datetime" column and save in separate column as integer vector named "hour"
taxi_data <- taxi_data %>% 
    mutate(hour = hour(pickup_datetime))

#Use Cross-join to create Data Frame with 24 hour intervals for each driver
hour <- 0:23
licence_crossed <- crossing(taxi_data$hack_license, hour) 
licence_crossed$hour <- as.numeric(as.character(licence_crossed $hour))
colnames(licence_crossed ) <- c("hack_license", "hour")


# Create Summary Dataframe of Total Passenger Count grouped by driver and hour
summary_total_passengers_picked_up <- taxi_data%>%
    filter(!is.na(passenger_count)) %>%
    group_by(hack_license, hour ) %>%
     summarize(total_passengers_picked_up = sum(passenger_count))

     


# Get Summary of Trips Started Stared By Hour
summary_trips_started <- taxi_data %>%
    filter(!is.na(pickup_datetime)) %>%
    group_by(hack_license, hour ) %>%
     summarize(trips_started= n())


#Join Summary Datasets together 
part_f_summary <- full_join(licence_crossed , summary_total_passengers_picked_up) #Join cross data with 
part_f_summary$total_passengers_picked_up <- replace(part_f_summary$total_passengers_picked_up, is.na(part_f_summary$total_passengers_picked_up), 0) 
part_f_summary <- full_join(part_f_summary, summary_trips_started) 
part_f_summary$trips_started <- replace(part_f_summary$trips_started, is.na(part_f_summary$trips_started), 0)
part_f_summary


'The summary dataframe below only includes information from trips where the hour of pickup was known.
Also there were rows that included pickup times but not a passenger count.  Consequently,
dividing by passegeners by trips_started to get an average rate using the table below would be misleading.'

```


##Part G
```{r}

'5 cases for each row 
     hour_interval > hour(pickup_time) & hour_interval = hour(dropoff_time) 
     hour_interval == hour(pickup_time) &   hour_interval == hour(dropoff_time) 
     hour_interval == hour(pickup_time) &   hour_interval < hour(dropoff_time) 
    hour_interval < hour(pickup_time) &   hour_interval > hour(dropoff_time) 
     ELSE 0 
'

'Not sure how to deal with NAs'
















```


